Complementing theory on adolescent emotion regulation using  machine learning

### 2022-02-27 manuscript/manuscript.Rmd
update 2022-02-27 11:00:23

Trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
The full sample fit was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
As this model is relatively complex for the subsequent SEM-forests analysis, residual variances were constrained over time, reflecting an assumption of homoscedasticity,
and variances of growth parameters were fixed at zero,
reflecting the assumption of homogeneity of trajectories within terminal nodes,
with remaining heterogeneity attributed to error variance.

To identify predictors of these latent growth trajectories, we used the SEM forests algorithm [@brandmaierTheoryguidedExplorationStructural2016].
This algorithm proceeds as follows:

@@

This study used the machine learning algorithm SEM-forests
to identify the most important early indicators of emotion regulation development across adolescence,
and to explore the nature of the association of those predictors with development. 
The results identified several potential protective- and risk factors.
Although all predictors attained a positive variable importance value,
there were differences in relative importance.
Individual differences in personality, motivation, self-concept clarity, and personal distress were among the most important predictors.
Conflict frequency and behavior also ranked highly.
Parenting ranked relatively low, with two notable exceptions:
First, balanced relatedness, a measure of autonomy support, was an important protective factor, and second,
among all remaining parenting constructs, negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
Other themes that ranked very low were family demographics,
bullying, victimization and delinquency, and substance use.

Relating these results to prior theory,
it is noteworthy that the order of importance of these themes appears to correspond to the levels of the bioecological model.
Predictors at the individual level were most important,
followed by those at the microsystem level, such as aspects of the relationship with parents,
followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status,
and the macrosystem indicator religiosity was among the least important predictors.

The present findings have important implications for intervention.
However, it is important to consider the correlational nature of the study design.
Our findings reveal the strength of the association of each predictor with emotion regulation development.
Intervention can only be effective if this association can additionally be assumed to be causal, based on theory or experimental research.
Two phenomena appear to be prime candidates for intervention,
based on the variable importance rankings:
conflict resolution behavior, and (negative) parenting.

There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.

Parenting behavior is another promising target for intervention,
as indeed it has historically been.
Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
By contrast, promoting positive parenting practices may have limited potential for impact.
With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
another meta-analysis called these findings into question,
citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
<!-- Another study Even when negative parenting has been assessed as outcomes, -->
<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
the ultimate benefit for children,
in particular for adolescents' emotion regulation,
is less studied.
The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
that negative stimuli are more psychologically impactful than positive ones,
which has also been found in children [@vaishNotAllEmotions2008].
Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices.

As parents have a legal and moral imperative to care for their children,
much of developmental research has understandably focused on parents' role in child and adolescent development.
It is thus noteworthy that, with the exception of autonomy support and negative parenting,
other parenting behaviors were rated low in terms of predictive importance.
This finding seems paradoxical in light of prior theory and empirical research.
At the same time, these findings are consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence.
At this point, the evidence suggests that we must consider the possibility that parents' influence in adolescence is more limited than once was thought.
Based on our findings, the best advice to give parents might be to support their individuating adolescents' autonomy,
refrain from overly invasive or negative parenting practices,
and model constructive problem solving behavior when conflicts arise.

## Reflection content

@@

@@

@@

It is also noteworthy that some predictors had relatively high predictive importance,
but almost no marginal effect (e.g., father's age, drug use, openness to experience).
Such predictors likely derive their importance from interactions.
For instance, drug use might be a vehicle for social status or self-harm;
older fathers might be an asset if they are involved in children's lives but a liability if they prioritize professional success;
openness to experience may be an asset if it leads to self-discovery but a liability if it leads to risk behavior.
As the present method is not well-suited to identify what interaction drives a variable's importance,
these results should be taken as indications for potentially fruitful areas of future research.

## Limitations

* Selective sample
  + WEIRD, Dutch, mostly mid- high-SES (restriction of range)
* Panel study, so few biological / cognitive / neurological measures
* Relatively small sample size
  + If you have access to 1000+ sample, please get in touch
* Computational constraints prevented us from tuning the model beyond default settings, and from exploring marginal effects for higher order interactions

@@

* Proximal factors more predictive of ER development than distal predictors
* Parenting may be less relevant than previously thought
* Personality and conflict behavior are candidates for early diagnosis
* Conflict skills training might be avenue for intervention

>>>>

Preliminary analyses were conducted to ensure the appropriateness of growth curve modeling.
To this end, trajectories of emotion regulation development were modeled using a curvilinear latent growth curve model with an intercept (I), linear slope (S), and quadratic slope (Q).
The full sample fit of this model was excellent, $\text{RMSEA} < .05, \text{SRMR} = 0.02, \text{CFI} = 1.00$.
Second, latent class growth analyses were conducted to determine the presence of youth with trajectories that indicated elevated risk of emotion regulation difficulties.
We used the R-package `tidySEM` to estimate and compare 1-6 class solutions of the aforementioned curvilinear growth curve model;
analysis results are displayed in [Supplemental Table 2]().
To ensure model convergence in small classes, the latent variable covariance matrix was restricted to zero,
reflecting the assumption of homogeneity of trajectories,
with remaining heterogeneity attributed to error variance.
Class solutions with more than two classes were excluded from consideration due to low class separability (entropy < .9).
The LMR likelihood ratio test indicated that two classes significantly improved fit ($p < .001$); more classes did not ($p = .26$).
As displayed in Figure \@ref(fig:figclass),
this analysis indicated a similar proportion of youth at-risk for difficulties in emotion regulation to what was found in previous studies [cf @leeAdolescentMentalHealth2014].

```{r figclass, fig.cap="Latent class analysis of risk profiles."}
knitr::include_graphics("../classplot.pdf")
```

In latent class analysis, individuals are clustered only with respect the the outcome (growth trajectories), to ensure maximum homogeneity of growth trajectories in each class.
It thus answers the question: "Which adolescents have similar growth trajectories?"
SEM-forests similarly aim to ensure maximum homogeneity of growth trajectories within clusters,
but they differ from latent class analysis in that they cluster individuals by reference to a set of predictors.
They thus answer the question: "Why do some adolescents have similar growth trajectories"?

For the SEM-forests analysis,
we imposed an additional restriction on the diagonal of the residual covariance matrix, reflecting an assumption of homoscedastic residual variance over time.
The SEM-forests algorithm operates as follows:

@@

This study used the machine learning algorithm SEM-forests
to identify the most important early indicators of emotion regulation development across adolescence,
and to explore the nature of the association of those predictors with development. 
The results identified several potential protective- and risk factors.
Although all predictors attained a positive variable importance value,
there were differences in relative importance.
Individual differences in personality, motivation, self-concept clarity, and personal distress were among the most important predictors.
Conflict frequency and behavior also ranked highly.
Parenting ranked relatively low, with two notable exceptions:
First, balanced relatedness, a measure of autonomy support, was an important protective factor, and second,
among all remaining parenting constructs, negative and intrusive dimensions were more predictive (as risk factors) than positive or neutral parenting practices.
Themes that ranked notably low were family demographics (SES and religion),
bullying, victimization and delinquency, and substance use.

## Implications for theory

Relating these results to prior theory,
it is noteworthy that predictive importance of different themes appears to correspond to the levels of the bioecological model:
Predictors at the individual level were most important,
followed by those at the microsystem level, such as aspects of the relationship with parents,
followed by the mesosystem, including aspects of the peer network (e.g., bullying/victimization and delinquency), school performance, and socio-economic status,
and the macrosystem indicator religiosity was among the least important predictors.

@@

As parents have a legal and moral imperative to care for their children,
much of developmental research has understandably focused on parents' role in child and adolescent development.
It is thus noteworthy that, with the exception of autonomy support and negative parenting,
other parenting behaviors were rated low in terms of predictive importance.
This finding seems paradoxical in light of prior theory and empirical research.
At the same time, these findings are consistent with an increasing number of recent studies - particularly those employing innovative methods to study processes within families - that have found little evidence for parent-directed processes in adolescence.
Of course, an alternative explanation might be that parenting effects take place earlier in development,
and are engrained in individual differences by the time youth reach adolescence.
A recent study on the role of parents in younger children's emotion regulation development casts doubt on this hypothesis however, as it similarly found little evidence for parent-directed processes.
<!-- aren’t individual differences the result of the environment where those people were growing up, hence parental style as well? -->
We must thus consider the possibility that parents' influence, at least in adolescence, is less prominent than once was thought.
Based on present findings, the best advice to give parents might be to support their individuating adolescents' autonomy,
refrain from overly invasive or negative parenting practices,
and model problem solving behavior when conflicts arise.

@@

## Implications for practice

The present findings also have important implications for intervention.
However, it is important to consider the correlational nature of the study design.
Our findings reveal the strength of the association of each predictor with emotion regulation development.
Intervention can only be effective if this association can additionally be assumed to be causal, based on theory or experimental research.
Two phenomena appear to be prime candidates for intervention,
based on the variable importance rankings:
conflict resolution behavior, and (negative) parenting.

There is prior evidence for the efficacy of conflict resolution training in adolescents [@breunlinConflictResolutionTraining2002].
The same evidence indicates that conflict resolution training caused a decrease in externalizing problems (another important predictor), suggesting potential compounding benefits.
Parenting behavior is another promising target for intervention,
as indeed it has historically been.
Importantly, our results suggest that such interventions should focus on promoting parental autonomy support, and decreasing intrusive, negative, guilt-inducing and overbearing parenting behaviors.
By contrast, promoting positive parenting practices may have limited potential for impact.
With this in mind, it is important to consider that positive parenting practices are often a key focus of parenting training.
Although one meta-analysis did report large effects of the Triple P-Positive Parenting Program on parental outcomes - including some negative parenting behaviors [@fletcherImpactBehaviouralParent2011],
another meta-analysis called these findings into question,
citing mixed evidence, a lack of adequate blinding, risk of bias, and the substantial cost of such programs [@wilsonHowEvidencebasedEvidencebased2012].
<!-- Another study Even when negative parenting has been assessed as outcomes, -->
<!-- significant effects were found for positive practices only [@marcynyszynGettingEvidencebasedProgram2011]. -->
<!-- and that there have been prior concerns about the efficacy of such training [@wilsonHowEvidencebasedEvidencebased2012; @cottamDiscoursesUnderpinningParenting2014]. -->
Also note that the efficacy of parenting training has been evaluated primarily in terms of parents' self-reported outcomes;
the ultimate benefit for children,
in particular for adolescents' emotion regulation,
is less studied.
The relatively higher importance rankings for negative versus parenting is in line with the well-established *negativity bias*,
that negative stimuli are more psychologically impactful than positive ones,
which has also been found in children [@vaishNotAllEmotions2008].
Future parenting interventions might take this into consideration by devoting sufficient attention to the elimination of negative parenting practices,
and including children's outcomes, including emotion regulation, into consideration.

## Methodological implications

@@

It is also noteworthy that some predictors had relatively high predictive importance,
but almost no marginal effect (e.g., father's age, drug use, openness to experience).
Such predictors likely derive their importance from interactions.
For instance, drug use might be a vehicle for social status or self-harm;
older fathers might be an asset if they are involved in children's lives but a liability if they prioritize professional success;
openness to experience may be an asset if it leads to self-discovery but a liability if it leads to risk behavior.
The present method is not well-suited to identify what interaction drives a variable's importance,
but these results might indicate potentially fruitful areas of future research.

## Strengths and limitations

The present study had some unique strengths, most notably that it constituted the first comprehensive study of potential predictors of adolescent emotion regulation development, where prior work has taken more piecewise approaches, considering only a few predictors at a time.
<!-- The present study set out to identify the most important early indicators of developmental trajectories -->
<!-- of emotion regulation across adolescence. -->
<!-- This involves three important challenges: -->
<!-- First, the number of potentially relevant risk factors, as identified in a prior text mining systematic review, is large. -->
<!-- Second, there is a paucity of relevant theory to narrow down the candidate predictors. -->
<!-- Third, methods commonly used in the field are not well suited for large numbers of predictors. -->
<!-- Machine learning methods that perform variable selection can overcome these challenges. -->
<!-- We therefore used the SEM-forests algorithm to identify the most important early predictors of trajectories of emotion regulation development throughout adolescence. -->
<!-- The primary aim is to identify the most important predictors of adolescents' trajectories of emotion regulation development. -->
<!-- The secondary aim is to probe the nature of the marginal association of the predictors with these trajectories. -->

The study also has some limitations, most notably that the sample was selective,
which may limit the generalizability of the findings.
Specifically, this was a non-random, ethnically and socially relatively homogenous  sample from the Netherlands.
This raises some specific concerns, for example, that the low importance of SES might be due to a restriction of range.
The sample might also be somewhat dated - especially considering daily life has changed drastically for youth coming of age in a post-pandemic world.
The only way to address these concerns is to replicate these findings in other samples.
Note that all code is available in a reproducible archive, which lowers the threshold to replication.
Another limitation is that the data source was a panel study with mostly questionnaire measures and relatively few biological, cognitive and neurological assays.
One final limitation related to the data is the relatively small sample size.
Machine learning methods benefit from large samples.
Although random forests are relatively robust to overfitting,
larger samples will likely yield more generalizable results,
and also afford more complicated models (thus removing the need for assumptions such as homogeneity of residuals).
Other limitations pertain to the method; for instance, SEM-forests require vast computational resources.
Computational constraints thus prevented us from tuning the model beyond default settings,
and from exploring marginal effects for higher order interactions.
Moreover, at the time of writing, there is not yet a method to evaluate the overall fit of a SEM-forest.

@@

Empirical research on adolescent emotion regulation development has been fragmented,
and lacks an overarching theoretical framework. 
The present study took a holistic approach to identify the most important predictors of emotion regulation development using machine learning.
Candidate predictors were selected based on a prior systematic review of 6305 relevant papers.
The SEM-forest algorithm was applied to rank-order these predictors by their relative importance for explaining heterogeneity in adolescents' developmental trajectories,
and to probe the nature of their association with emotion regulation development. 
Results indicated that the importance of different predictors corresponded somewhat with the levels of the bioecological model:
Proximal factors were more predictive of emotion regulation development than distal predictors; individual differences were most predictive, and mesosystem factors (e.g., SES, religion) were least predictive.
The results also suggested that parenting may be less relevant for emotion regulation development than previously thought, and that negative parenting may be more impactful than positive parenting.
The results have clear implications for early risk assessment:
Regularly administered personality inventories, and readily observable indicators like externalizing problems and conflict behavior were among the most important predictors, and these are suitable indicators for diagnostic tools.
The results further suggest that conflict skills training might be a promising avenue for adolescent-directed interventions.
Parenting interventions might benefit more from addressing the notion of "good enough" parenting and remedying negative and invasive behaviors than from promoting positive parenting behaviors.
The results also provide clear guidance for the ongoing effort of theory generation in this field:
Such theory should focus on proximal processes over distal ones.
Parents' roles might be less central than previously thought;
their support of adolescents' autonomy and the way they navigate conflict might be more important than their parenting behaviors.
Moreover, there was no evidence of differences between mothers' and fathers' parenting, indicating that both parents should be equally represented in future theory.
Finally, common perceptions about risk factors should be reassessed: 
delinquency, substance use, and bullying were all relatively less important predictors.
Even drug use, the highest-ranking problem behavior, had no distinguishable marginal effect, suggesting that it is more problematic for some adolescents than others.
These results illustrate how machine learning can provide a holistic view of a research area,
complement existing knowledge,
offer alternative perspectives,
and guide future research, practice, and theory formation.
